Inference of a universal social scale and segregation measures using social connectivity kernels

### 2020-10-05 manuscript/main.tex
Split document into main and supplement.

\documentclass{scrartcl}
\usepackage{authblk}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{csquotes}
\usepackage[british]{babel}
\usepackage[
    giveninits,
    backend=biber,
    sorting=none,
    citestyle=numeric-comp,
    maxnames=10,
    style=vancouver,
]{biblatex}
\usepackage{hyperref}
\usepackage{tabularx}
\usepackage{booktabs}
\usepackage{cleveref}
\usepackage{fmtcount}
\usepackage{environ}
\usepackage{graphicx}
\DeclareGraphicsExtensions{.pdf}

\usepackage[usenames]{xcolor}
\definecolor{mpl0}{HTML}{1f77b4}
\definecolor{mpl1}{HTML}{ff7f0e}
\definecolor{mpl2}{HTML}{2ca02c}
\definecolor{mpl3}{HTML}{d62728}
\definecolor{mpl4}{HTML}{9467bd}
\definecolor{mpl5}{HTML}{8c564b}
\definecolor{mpl6}{HTML}{e377c2}
\definecolor{mpl7}{HTML}{7f7f7f}
\definecolor{mpl8}{HTML}{bcbd22}
\definecolor{mpl9}{HTML}{17becf}

@@

\newcommand{\expit}{\sigma}
\newcommand{\transpose}[1]{#1^\intercal}
\newcommand{\observed}[1]{#1}
\DeclareMathOperator{\logit}{logit}
\DeclareMathOperator{\logp}{log1p}
\DeclareMathOperator{\logsumexp}{logsumexp}

\newcommand{\abs}[1]{\left|#1\right|}
\newcommand{\card}[1]{\left|#1\right|}
\newcommand{\population}{N}
\newcommand{\seeds}{U}
\newcommand{\nominees}{V}
\newcommand{\missing}{R}
\newcommand{\blauspace}{\mathbb{B}}
\newcommand{\twocol}[1]{\multicolumn{2}{>{\hsize=\dimexpr2\hsize+3\tabcolsep+2\arrayrulewidth\relax}X}{#1}}

@@

\def\composite{}
% Useful for word counting
\ifx\composite\undefined
  % hide graphics
  \renewcommand{\includegraphics}[1]{}
  % hide pagenumber
  \pagestyle{empty}
  \renewcommand*{\titlepagestyle}{empty}
  % hide equations
  \RenewEnviron{equation}{equation environment}
  \RenewEnviron{align}{align environment}
  \RenewEnviron{equation*}{equation* environment}
\fi

@@

\begin{refsection}
\section{Introduction\label{sec:introduction}}

@@

\emph{Social} statistics of segregation, such as the assortativity coefficient~\cite{Newman2003a}, overcome these limitations by explicitly considering the interactions amongst individuals~\cite{Blumenstock2013}, but have their own difficulties: first, they usually rely on the existence of discrete groups, such as sex, ethnicity, or religion~\cite{Bojanowski2014}, and they are not applicable to continuous attributes, such as age or income. Attributes are often discretised~\cite{Lam-Morgan2012, Kalmijn2007, Kim2012}, but the boundaries between categories always suffer from some degree of arbitrariness~\cite{Reardon2004}. Second, segregation for multiple attributes can be quantified independently, but it is unclear how to define a composite segregation statistic. Furthermore, social statistics are typically defined as summary statistics of a fully-observed social network. Consequently, we cannot easily quantify uncertainties.

@@

A single statistic cannot capture the complexities of social networks, and we develop a family of statistics applicable at different scales: the social separation between any two individuals, the isolation experienced by any one individual, and the social strain experienced by society as a whole. Starting at the microscopic level, we use the generative model to define a statistic of \emph{social separation} between two individuals with attributes $x$ and $y$:
\begin{align}
    \varphi(x, y) &= \logit \rho(y,y) - \logit\rho(x,y),\label{eq:social-separation}\\
    \text{where }\logit\rho &= \log\left(\frac{\rho}{1-\rho}\right)\nonumber
\end{align}
are the log odds for a connection to form with probability $\rho$. The probability $\rho(y, y)$ for an individual with attributes $y$ to connect with someone with identical attributes serves as a reference point, and the statistic does not depend on the overall edge density. The social separation $\varphi$ may be understood as the isolation experienced by an individual with attributes $x$ as a result of the behaviour of another with attributes $y$. The statistic is zero if people do not discriminate with respect to the attributes on which they differ or if two individuals have the same demographic attributes. For a homophilous connectivity kernel, the statistic is positive and is a \emph{semi-metric} for Blau space; we will consider a family of connectivity kernels for which $\varphi$ is a true metric in \cref{sec:segregation-distance}.

@@

The connectivity kernel $\rho(x, y)$ is not sufficient to quantify segregation at the level of an individual: we also need to consider the distribution $P(y)$ of attributes $y$ of all other members of society, such as their age, sex, or other demographics. For an individual with attributes $x$, we define the \emph{social isolation}
\begin{equation}
    \phi(x) = \int dy \ P(y) \varphi(x, y),\label{eq:segregation-individual}
\end{equation}
which quantifies the average social separation between an individual with attribute $x$ and other members of society. For the SBM, we substitute \cref{eq:social-separation-sbm} into \cref{eq:segregation-individual} and obtain
\begin{equation}
    \phi(x) = (1-P(x))\left(\logit\rho_\mathrm{same}-\logit\rho_\mathrm{different}\right),\label{eq:segregation-individual-sbm}
\end{equation}
where $P(x)$ is the probability to belong to block $x$, and we have used the identity $\sum_{y=1}^K P(y)\left(1-\delta_{xy}\right)= 1 - P(x)$. Members of all blocks experience the same degree of isolation if the blocks are of the same size. If the sizes are unequal, minorities experience more isolation and majority groups experience less isolation. Indeed, ethnic minorities in schools tend to be more isolated and have fewer social ties~\cite{Currarini2009}.

@@

The social separation takes a simple form if the probability for two individuals to connect is a logistic kernel~\cite{Hoff2002}, i.e.\
\begin{equation}
    \logit\rho(x, y, \theta) = \sum_{l=1}^p\theta_l f_l(x,y),\label{eq:logistic-kernel}
\end{equation}
where the $p$-dimensional vector $\theta_l$ parametrises the kernel, and $f(x, y)$ is a set of $p$-dimensional features that are predictive of the connection probability, such as the age difference $f_\text{age}=\abs{x_\mathrm{age}-y_\mathrm{age}}$. The social separation between $x$ and $y$ comprises contributions from the features of the logistic kernel:
\begin{align}
    \varphi(x,y) &= \sum_{l=1}^p\varphi_l(x, y),\label{eq:social-separation-logistic}\\
    \text{where }\varphi_l(x, y) &= \theta_l\left(f_l(y, y) - f_l(x, y)\right)\label{eq:social-separation-contribution}
\end{align}
is the contribution due to a single feature $l$. In fact, $\varphi$ is a true metric for many logistic connectivity kernels.

@@

A representative sample of dyads between individuals together with their demographic attributes is not generally available. However, a number of surveys have collected information about the social ties of respondents using name-generator questions which elicit social ties by asking respondents to nominate their friends~\cite{Kalmijn2007}, individuals they feel close to~\cite{Hipp2009}, or discussion partners~\cite{Marsden1987,McPherson2006}. To generate examples of disconnected dyads, we consider a random sample of pairs of individuals. To account for this non-ignorable data collection process, we introduce a variable $I_{ij}\in\{0,1\}$ indicating whether a particular dyad $A_{ij}$ was observed. The available data thus comprise the demographic attributes $x$ of individuals included in the sample and the dyad state $A_{ij}$ (1 if $i$ and $j$ are connected and $0$ otherwise) if it was observed, i.e.\ $I_{ij}=1$. Adapting the argument presented by \textcite{King2001} to a Bayesian paradigm, we consider the posterior distribution over kernel parameters $\theta$ given the available data:
\begin{equation}
    P(\theta|\observed{A}, \observed{f}, I=1)\propto P(\observed{A}|\theta,\observed{f},I=1)P(\theta),\label{eq:parameter-posterior}
\end{equation}
where $P(\theta)$ is the kernel parameter prior, and $f=f(x,y)$ are features sufficient to evaluate the connectivity kernel given demographic attributes $x$ and $y$. The observed-data likelihood is
\begin{equation}
    P(\observed{A}|\theta,\observed{f},I=1)=\frac{P(\observed{f}|\observed{A},\theta,I=1)P(\observed{A}|\theta,I=1)}{P(\observed{f}|\theta,I=1)}.\label{eq:observed-likelihood}
\end{equation}
Considering the first term in the numerator of \cref{eq:observed-likelihood}, we note that the distribution over kernel features given the state $A$ of the dyad does not depend on whether it was included in the sample or not. More formally,
\begin{align}
    P(f|A,\theta,I=1)&=P(f|A,\theta)\\
    &=\frac{P(A|f,\theta)P(f|\theta)}{P(A|\theta)}.\label{eq:conditional-equivalence}
\end{align}
Turning to the denominator in \cref{eq:observed-likelihood}, we find
\begin{align}
    P(f|\theta,I=1)&=\sum_{\alpha=0}^1 P(f|A=\alpha,\theta,I=1)P(A=\alpha|\theta,I=1)\nonumber\\
    &=P(f|\theta)\sum_{\alpha=0}^1 P(A=\alpha|f,\theta)\frac{P(A=\alpha|\theta,I=1)}{P(A=\alpha|\theta)},\label{eq:likelihood-denominator}
\end{align}
where we used the identity in \cref{eq:conditional-equivalence} to arrive at the second line. Substituting \cref{eq:conditional-equivalence,eq:likelihood-denominator} into \cref{eq:observed-likelihood}, the observed-data likelihood is
\begin{align}
    P(A|\theta,f,I=1)&=\frac{P(A|f,\theta)r(A)}{\sum_{\alpha=0}^1 P(A=\alpha|f,\theta)r(\alpha)},\label{eq:observed-likelihood-final}\\
    \text{where }r(\alpha)&=\frac{P(A=\alpha|\theta,I=1)}{P(A=\alpha|\theta)}\nonumber
\end{align}
is the ratio of prevalences of dyad state $\alpha$ in the sample and the general population. In practice, we approximate the prevalence ratio $r$ using the empirical sample prevalence and prior knowledge about the prevalence in the population. The posterior can be evaluated by substituting \cref{eq:observed-likelihood-final} into \cref{eq:parameter-posterior}, and we can thus infer the parameters $\theta$ from ego network data. See \cref{app:numeric-stability} for details on how to evaluate the observed-data log-likelihood in a numerically stable fashion and \cref{app:inference-validation} for a validation of the inference methodology using synthetic data. For logistic connectivity kernels, the observed-data likelihood in \cref{eq:observed-likelihood-final} resembles a conventional case-control likelihood, e.g.\ as used by \textcite{Smith2014}.

@@

We considered a generative model for social networks embedded in Blau space, a space spanned by the demographic attributes of members of society. We developed a family of segregation statistics with a universal scale (since they are based on the common notion of the probability of a social tie) facilitating comparison between datasets collected at different times or in different cultural contexts. Furthermore, the segregation statistics are applicable to mixed attribute types, have a natural reference point, and an intuitive interpretation: the probability to form connections decreases with increasing segregation. For certain logistic connectivity kernels, the social separation is a metric for Blau space and allows us to quantify social distance in a principled fashion.

@@

The importance of physical distance highlights that our suite of segregation statistics does not distinguish between \emph{choice homophily} and \emph{opportunistic homophily}~\cite{Franz2010}. The former is a result of individuals having an active preference to connect with others who are alike, whereas the latter is a result of individuals being exposed to others who are similar to them. Opportunistic homophily is likely to be a large contributing factor to spatial homophily because individuals are less likely to encounter people who live far from them.

@@

The connectivity kernel is an intuitive model of how people connect with one another, and it is able to reproduce some of the statistics of real social networks. For example, people in high-density regions of Blau space have been observed to have more connections~\cite{Currarini2009}. However, exponential random graph models may be able to better capture the nature of social networks~\cite{Wimmer2010}. Furthermore, we have used a connectivity kernel that: (a)~is symmetric and cannot identify whether there is a status order in society~\cite{Chan2004,Ball2013}, and (b)~only depends on differences between individuals. For example, young men tend to have more social contacts than young women, and older women have more social contacts than older men~\cite{Bhattacharya2016}---an observation that cannot be captured by a kernel of the form we have considered. The connectivity kernel could be refined by adding the demographic attributes of the seeds and nominees as features, capturing sociability and popularity, respectively. Furthermore, it should be determined whether the number of ``intervening opportunities''~\cite{Stouffer1940}, absolute distance in Blau space, or a hybrid thereof are most predictive of tie probability. Ultimately, learning a connectivity kernel without a pre-specified parametric form should be considered~\cite{Frolich2006}. We note that, irrespective of the choice of connectivity kernel, the interpretable segregation statistics considered here remain valid and useful.

@@

We would like to thank Sahil Loomba for useful feedback on the manuscript and acknowledge support from the grant EP/N014529/1.

\printbibliography
\end{refsection}

\ifdefined\composite
\pagebreak
\noindent{\titlefont\huge Supplementary material}

\appendix
\numberwithin{equation}{section}
\numberwithin{figure}{section}
\numberwithin{table}{section}

\begin{refsection}
\section{Evaluation of the observed-data log-likelihood}

\subsection{Weighting to account for non-uniform inclusion probabilities}

Seeds are often not included in the survey uniformly at random, and weights are traditionally used to compensate for the potentially biased selection of respondents~\cite{Kish1992}. Including weights in Bayesian analyses is generally difficult~\cite{Gelman2007}, and, in principle, we should model the data collection process explicitly~\cite[chapter~8]{Gelman2013}. Unfortunately, modelling the data collection process is non-trivial, and we use a weighted pseudo-likelihood instead~\cite{Pfeffermann1996}. In particular, the observed-data log-likelihood from \cref{eq:observed-likelihood-final} becomes
\begin{multline}
    L = \sum_{(i,j):I_{ij} = 1} w_j\left\{A_{ij} + (1 - A_{ij})w_i\right\}
    \\\times\left\{A_{ij}\log\rho_{ij} + (1-A_{ij})\log(1-\rho_{ij})-\log\left[r(0)(1 - \rho_{ij}) + r(1)\rho_{ij}\right]\right\},\label{eq:survey-weighted-log-likelihood}
\end{multline}
where $w_j$ is the weight associated with seed $j$. We clip all weights exceeding the \ordinalnum{95} percentile of the empirical weight distribution and normalise them such that $\sum_{j\in\seeds}w_j=\card{\seeds}$. Censoring the weights, also known as Winsorisation, limits the variance induced by attributing variable importance to different observations at the expense of introducing a small bias~\cite{Kish1992}.

\subsection{Numerical stability\label{app:numeric-stability}}

The evaluation of the observed-data log-likelihood may suffer from numerical instabilities, especially when the connectivity kernel $\rho(x, y, \theta)$ is small. We can mitigate such instabilities for logistic connectiviy kernels, i.e.
\begin{align}
    \rho(x,y,\theta)&=\expit(\transpose{\theta}f(x,y)),\\
    \text{where }\expit(\xi)&=\frac{1}{1+\exp(-\xi)}
\end{align}
is the logistic function. In particular, note that $1 - \expit(\xi) = \expit(-\xi)$ and $\log\expit(\xi)=-\logp\left[\exp(-\xi)\right]$, where $\logp(\xi)=\log(1 + \xi)$ is a numerically stable implementation---even for $\abs{\xi}\ll 1$. Substituting into \cref{eq:observed-likelihood-final} yields
\begin{multline}
    \log P(A|f,\theta,I=1)= -\sum_{(i,j):I_{ij}=1} A_{ij}\logp\exp\left(-\transpose{\theta}f_{ij}\right)+\left(1-A_{ij}\right)\logp\exp\left(\transpose{\theta}f_{ij}\right) \\
    + \logsumexp\left[\log r(0) - \logp\exp\left(\transpose{\theta}f_{ij}\right) + ,\log r(1)-\logp\exp\left(-\transpose{\theta}f_{ij}\right)\right],
\end{multline}
where $\logsumexp(x_1, \ldots, x_k)=\log\sum_{i=1}^k\exp(x_k)$ is a numerically stable implementation.

\section{Validation of inference methodology using synthetic ego network data\label{app:inference-validation}}

\begin{figure}
    \includegraphics{credible-coverage}
    \titlecaption{A coverage analysis of posterior credible intervals validates the inference methodology.}{The blue line shows the fraction of inferences for which the true parameter values are contained in the $\alpha$-credible interval of a Laplace approximation of the posterior. The shaded region corresponds to two standard deviations of the mean across 250 simulations.\label{fig:credible-coverage}}
\end{figure}

To test the inference methodology, we conduct a coverage analysis of posterior credible intervals in three steps: fist, we generate 250 synthetic ego network datasets with known kernel parameter values. Second, we infer the parameter posterior distribution. Third, we evaluate the proportion $\lambda(\alpha)$ of true parameter values contained in the $\alpha$-credible interval across multiple synthetic datasets. We expect the true parameter values to be contained in the $\alpha$-credible interval for a proportion $\alpha$ of the synthetic datasets~\cite[section~10.7]{Gelman2013}, i.e.\ $\lambda(\alpha)\approx\alpha$.

In the first step, we draw the positions $x$ of $n=2,000$ nodes uniformly at random from the unit square, and we connect nodes to one another according to a logistic connectivity kernel with features
\begin{equation*}
    f(x_i, x_j) = \left(1, \frac{3 \left[\abs{x_{i1}-x_{j1}} - \frac{1}{3}\right]}{\sqrt{2}}, \frac{3 \left[\abs{x_{i2}-x_{j2}} - \frac{1}{3}\right]}{\sqrt{2}}\right),
\end{equation*}
where the first feature represents the bias, and the last two features capture distance in Blau space. The features were chosen to be standardised in the same fashion as described in \cref{sec:survey-gss}, i.e.\ to have zero mean and a standard deviation of $0.5$~\cite{Gelman2008a}. The corresponding parameter values $\theta$ are drawn from a normal distribution with unit variance and expectation
\begin{equation*}
    \left\langle \theta\right\rangle = \left(-7, 0, 0\right).
\end{equation*}
The expectation $\langle\theta\rangle$ was chosen such that it gives rise to a typical degree of $2,000 \times \expit(-7)\approx 1.8$, similar to the real-world datasets considered in \cref{sec:application}. We select $s=100$ respondents as egos and include all their alters as positive examples. We sample three times as many negative examples by selecting distinct pairs of respondents. If the number of respondents is not sufficient to draw the desired number of distinct control pairs, we use all $\frac{s(s-1)}{2}$ possible distinct pairs of respondents as negative examples.

In the second step, we maximise the log-posterior using a gradient ascent algorithm to obtain the MAP estimate $\theta$ and consider the Laplace approximation of the posterior~\cite[section~4.4]{Bishop2007}, i.e.\ a multivariate normal approximation in the vicinity of the MAP estimate. We evaluate the Hessian $H$ of the negative log-posterior at the MAP estimate to obtain the precision matrix of the Laplace approximation. We do not draw samples from the posterior because the Laplace approximation is computationally more convenient.

In the last step, we consider the quantity
\begin{equation}
   \chi^2 = \transpose{\left(\theta-\hat\theta\right)}H\left(\theta-\hat\theta\right)\label{eq:survey-chi2-statistic}
\end{equation}
which we expect to follow a $\chi^2$ distribution with three degrees of freedom~\cite{Slotani1964}. We calculate the statistic in \cref{eq:survey-chi2-statistic} for each simulation and consider the empirical probability that $\chi^2$ does not exceed the expected quantiles of the $\chi^2$-distribution, as shown in \cref{fig:credible-coverage}. As expected, the $\alpha$-credible interval contains the true parameter values for a fraction $\alpha$ of inferences, validating the inference methodology.

\section{Coding of demographic attributes and feature maps\label{app:coding-and-features}}

In the BHPS and USS, distance was coded as an ordinal variable: less than one mile, less than five miles, less than fifty miles, and more than fifty miles. We rely on having complete information about seeds to evaluate the control features in \cref{eq:logistic-kernel}. But data on the residential location of seeds is not made available to protect their privacy. Fortunately, we can sample the home locations of respondents\footnote{Sampling home locations cannot reproduce any correlation between home location and other demographic attributes.} using population estimates and the geographic boundaries of lower layer super output areas (LSOAs). LSOAs are census reporting areas and have a few thousand inhabitants each~\cite{Lsoas}. We approximate the distribution of distances between residents of the UK using rejection sampling: first, choose a LSOA with probability proportional to the number of residents. Second, choose one of the polygons associated with the LSOA with probability proportional to the area of the polygon (LSOAs are not necessarily contiguous). Third, sample points uniformly inside the bounding box of the polygon until a point inside the polygon is sampled. The last two steps assume uniform population densities within each LSOA, which is unlikely to be problematic as they are small areas. Having sampled the residential location of two respondents, we calculate the distance between respondents and cast to the same ordinal scale as reported for nominees. The USS furthermore distinguishes between friends living more than fifty miles apart but within the UK and friends outside the UK. We discard the latter (2.6\% and 2.1\% of all friends in waves C and F of the USS) because it is difficult to define an appropriate control population. For the BHPS, we implicitly assume that all friends are resident in the UK.

In the USS, respondents could identify with mixed ethnicities, and we coded such responses as a mixed membership. For example, a respondent who indicated ``mixed Asian and White'' would belong to both White and Asian ethnicities. To quantify how different two people are in terms of ethnicity, we define the feature map
\begin{equation*}
    f_\text{ethnicity}(x_i, x_j) = \frac{1}{2} \sum_{l\in E}\abs{x_{il}-x_{jl}},
\end{equation*}
where $E$ is the set of attributes encoding ethnic identity, and ethnicity memberships are normalised such that $\sum_{l\in E}x_{jl}=1$ for all $j$. For example, $f_\text{ethnicity}(x_i, x_j)=1$ for two people $i$ and $j$ one of which identifies as white and the other as black. For a person $i$ identifying as black and another person $j$ identifying as mixed black and Asian, $f_\text{ethnicity}(x_i, x_j)=0.5$.

>>>>

\documentclass{scrartcl}
\usepackage{authblk}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{csquotes}
\usepackage[british]{babel}
\usepackage[
    giveninits,
    backend=biber,
    sorting=none,
    citestyle=numeric-comp,
    maxnames=10,
    style=vancouver,
]{biblatex}
\usepackage{xr-hyper}
\usepackage{hyperref}
\usepackage{tabularx}
\usepackage{booktabs}
\usepackage{cleveref}
\usepackage{fmtcount}
\usepackage{environ}
\usepackage{graphicx}
\DeclareGraphicsExtensions{.pdf}

\externaldocument{supplement}

\usepackage[usenames]{xcolor}
\definecolor{mpl0}{HTML}{1f77b4}

@@

\newcommand{\expit}{\sigma}
\newcommand{\transpose}[1]{#1^\intercal}
\DeclareMathOperator{\logit}{logit}
\DeclareMathOperator{\logp}{log1p}
\DeclareMathOperator{\logsumexp}{logsumexp}

\newcommand{\abs}[1]{\left|#1\right|}
\newcommand{\card}[1]{\left|#1\right|}
\newcommand{\population}{N}
\newcommand{\seeds}{U}
\newcommand{\blauspace}{\mathbb{B}}
\newcommand{\twocol}[1]{\multicolumn{2}{>{\hsize=\dimexpr2\hsize+3\tabcolsep+2\arrayrulewidth\relax}X}{#1}}

@@

@@

\section{Introduction\label{sec:introduction}}

@@

\emph{Social} statistics of segregation, such as the assortativity coefficient~\cite{Newman2003a}, overcome these limitations by explicitly considering the interactions amongst individuals~\cite{Blumenstock2013}, but have their own difficulties: first, they usually rely on the existence of discrete groups, such as sex, ethnicity, or religion~\cite{Bojanowski2014}, and they are not applicable to continuous attributes, such as age or income. Attributes are often discretised~\cite{Lam-Morgan2012, Kalmijn2007, Kim2012}, but the boundaries between categories always suffer from some degree of arbitrariness~\cite{Reardon2004}. Second, segregation for multiple attributes can be quantified independently, but \textcite{Pelechrinis2016}, who consider a multivariate generalisation of the assortativity coefficient, note that ``a formal metric that is generally applicable'' remains elusive. Furthermore, social statistics are typically defined as summary statistics of a fully-observed social network. Consequently, we cannot easily quantify uncertainties.

@@

A single statistic cannot capture the complexities of social networks, and we develop a family of statistics applicable at different scales: (a)~the \emph{social separation} between any two individuals, (b)~the \emph{isolation} experienced by any one individual, and (c)~the \emph{social strain} experienced by society as a whole. Starting at the microscopic level, we define the \emph{social separation} between two individuals $i$ and $j$ with attributes $x$ and $y$ as the relative log odds for $j$ to connect with $i$ compared someone who is alike: the log odds intuitively capture the probabilistic nature of the conditionally independent edge model. In particular,
\begin{align}
    \varphi(x, y) &= \logit \rho(y,y) - \logit\rho(x,y),\label{eq:social-separation}\\
    \text{where }\logit\rho &= \log\left(\frac{\rho}{1-\rho}\right)\nonumber
\end{align}
are the log odds for a connection to form with probability $\rho$~\cite{Hastie2009}. The probability $\rho(y, y)$ for $j$ to connect with someone who is alike serves as a reference point, and the statistic does not depend on the overall edge density. The social separation $\varphi$ may be understood as the isolation experienced by $i$ with attributes $x$ as a result of the behaviour of $j$ with attributes $y$. The statistic is zero if people do not discriminate with respect to the attributes on which they differ or if two individuals have the same demographic attributes. For a homophilous connectivity kernel, the statistic is positive and is a \emph{semi-metric} for Blau space; we will consider a family of connectivity kernels for which $\varphi$ is a true metric in \cref{sec:segregation-distance}.

@@

The connectivity kernel $\rho(x, y)$ is not sufficient to quantify segregation at the level of an individual: we also need to consider the distribution $P(y)$ of attributes $y$ of all other members of society, such as their age, sex, or other demographics. For an individual with attributes $x$, we define the \emph{social isolation}
\begin{equation}
    \phi(x) = \int dy \ P(y) \varphi(x, y),\label{eq:segregation-individual}
\end{equation}
which quantifies the average social separation between an individual with attribute $x$ and other members of society. For the SBM, we substitute \cref{eq:social-separation-sbm} into \cref{eq:segregation-individual} and obtain
\begin{equation}
    \phi(x) = (1-P(x))\left(\logit\rho_\mathrm{same}-\logit\rho_\mathrm{different}\right),\label{eq:segregation-individual-sbm}
\end{equation}
where $P(x)$ is the probability to belong to block $x$, and we have used the identity $\sum_{y=1}^K P(y)\left(1-\delta_{xy}\right)= 1 - P(x)$. Members of all blocks experience the same degree of isolation if the blocks are of the same size. If the sizes are unequal, minorities experience more isolation and majority groups experience less isolation. Indeed, ethnic minorities in schools tend to be more isolated and have fewer social ties~\cite{Currarini2009}. The off-diagonal terms of the Hessian of the social isolation $\phi$ quantify interactions, such as the joint effect of age and ethnic differences.

@@

The social separation takes a simple form if the probability for two individuals to connect is a logistic kernel~\cite{Hoff2002}, i.e.\
\begin{equation}
    \logit\rho(x, y, \theta) = \sum_{l=1}^p\theta_l f_l(x,y),\label{eq:logistic-kernel}
\end{equation}
where the $p$-dimensional vector $\theta_l$ parametrises the kernel, and $f(x, y)$ is a set of $p$-dimensional features that are predictive of the connection probability, such as the age difference $f_\text{age}=\abs{x_\mathrm{age}-y_\mathrm{age}}$. Intersectionality can be accounted for by including interaction terms in the feature set. The social separation between $x$ and $y$ comprises contributions from the features of the logistic kernel:
\begin{align}
    \varphi(x,y) &= \sum_{l=1}^p\varphi_l(x, y),\label{eq:social-separation-logistic}\\
    \text{where }\varphi_l(x, y) &= \theta_l\left(f_l(y, y) - f_l(x, y)\right)\label{eq:social-separation-contribution}
\end{align}
is the contribution due to a single feature $l$. In fact, $\varphi$ is a true metric for many logistic connectivity kernels.

@@

A representative sample of dyads between individuals together with their demographic attributes is not generally available. However, a number of surveys have collected information about the social ties of respondents using name-generator questions which elicit social ties by asking respondents to nominate their friends~\cite{Kalmijn2007}, individuals they feel close to~\cite{Hipp2009}, or discussion partners~\cite{Marsden1987,McPherson2006}. To generate examples of disconnected dyads, we consider a random sample of pairs of individuals. To account for this non-ignorable data collection process, we introduce a variable $I_{ij}\in\{0,1\}$ indicating whether a particular dyad $A_{ij}$ was observed. The available data thus comprise the demographic attributes $x$ of individuals included in the sample and the dyad state $A_{ij}$ (1 if $i$ and $j$ are connected and $0$ otherwise) if it was observed, i.e.\ $I_{ij}=1$. Adapting the argument presented by \textcite{King2001} to a Bayesian paradigm, we consider the posterior distribution over kernel parameters $\theta$ given the available data:
\begin{equation}
    P(\theta|A, f, I=1)\propto P(A|\theta,f,I=1)P(\theta),\label{eq:parameter-posterior}
\end{equation}
where $P(\theta)$ is the kernel parameter prior, and $f=f(x,y)$ are features sufficient to evaluate the connectivity kernel given demographic attributes $x$ and $y$. The observed-data likelihood is
\begin{equation}
    P(A|\theta,f,I=1)=\frac{P(f|A,\theta,I=1)P(A|\theta,I=1)}{P(f|\theta,I=1)}.\label{eq:observed-likelihood}
\end{equation}
Considering the first term in the numerator of \cref{eq:observed-likelihood}, we note that the distribution over kernel features given the state $A$ of the dyad does not depend on whether it was included in the sample or not. More formally,
\begin{align}
    P(f|A,\theta,I=1)&=P(f|A,\theta)\\
    &=\frac{P(A|f,\theta)P(f|\theta)}{P(A|\theta)}.\label{eq:conditional-equivalence}
\end{align}
Turning to the denominator in \cref{eq:observed-likelihood}, we find
\begin{align}
    P(f|\theta,I=1)&=\sum_{\alpha=0}^1 P(f|A=\alpha,\theta,I=1)P(A=\alpha|\theta,I=1)\nonumber\\
    &=P(f|\theta)\sum_{\alpha=0}^1 P(A=\alpha|f,\theta)\frac{P(A=\alpha|\theta,I=1)}{P(A=\alpha|\theta)},\label{eq:likelihood-denominator}
\end{align}
where we used the identity in \cref{eq:conditional-equivalence} to arrive at the second line. Substituting \cref{eq:conditional-equivalence,eq:likelihood-denominator} into \cref{eq:observed-likelihood}, the observed-data likelihood is
\begin{align}
    P(A|\theta,f,I=1)&=\frac{P(A|f,\theta)r(A)}{\sum_{\alpha=0}^1 P(A=\alpha|f,\theta)r(\alpha)},\label{eq:observed-likelihood-final}\\
    \text{where }r(\alpha)&=\frac{P(A=\alpha|\theta,I=1)}{P(A=\alpha|\theta)}\nonumber
\end{align}
is the ratio of prevalences of dyad state $\alpha$ in the sample and the general population. In practice, we approximate the prevalence ratio $r$ using the empirical sample prevalence and prior knowledge about the prevalence in the population. The posterior can be evaluated by substituting \cref{eq:observed-likelihood-final} into \cref{eq:parameter-posterior}, and we can thus infer the parameters $\theta$ from ego network data. See \cref{app:numeric-stability} for details on how to evaluate the observed-data log-likelihood in a numerically stable fashion and \cref{app:inference-validation} for a validation of the inference methodology using synthetic data. For logistic connectivity kernels, the observed-data likelihood in \cref{eq:observed-likelihood-final} resembles a conventional case-control likelihood, e.g.\ as used by \textcite{Smith2014}.

@@

We considered a generative model for social networks embedded in Blau space, a space spanned by the demographic attributes of members of society. We developed a family of segregation statistics with a universal scale (since they are based on the common notion of the probability of a social tie), facilitating comparison between datasets collected at different times or in different cultural contexts. Furthermore, the segregation statistics are applicable to mixed attribute types, have a natural reference point, and an intuitive interpretation: the probability to form connections decreases with increasing segregation. They are applicable at different resolutions: connections, individuals, and society as a whole. For certain logistic connectivity kernels, the social separation is a metric for Blau space and allows us to quantify social distance in a principled fashion. The model-based approach facilitates the study of segregation in synthetic social networks, the effect of interventions, and principled quantification of uncertainties in an applied setting.

@@

The importance of physical distance highlights that our suite of segregation statistics does not distinguish between \emph{choice homophily} and \emph{opportunistic homophily}~\cite{Franz2010}. The former is a result of individuals having an active preference to connect with others who are alike, whereas the latter is a result of individuals being exposed to others who are similar to them. Opportunistic homophily is likely to be a large contributing factor to spatial homophily because individuals are less likely to encounter people who live far from them. Similarly, the statistics do not discriminate between choice homophily and \emph{social influence}, i.e.\ the tendency for people to become more alike given a connection~\cite{Shalizi2011}.

Other features, including sex, ethnicity, religion, education, and occupation, have smaller effects on the presence of connections. Notably, the social strain due to ethnicity in the GSS and ALP is larger than in the USS: first, the effect of ethnicity is more pronounced in the US, as shown in \cref{fig:parameters-and-age-equivalents}. Second, society in the US is more ethnically diverse than in the UK (79\% white in GSS '04 compared with 89\% white in USS '14), increasing strain on average, as exemplified with a SBM in \cref{sec:model-based-segregation}.

@@

The connectivity kernel is an intuitive model of how people connect with one another, and it is able to reproduce some of the statistics of real social networks. For example, people in high-density regions of Blau space have been observed to have more connections~\cite{Currarini2009}. However, exponential random graph models may be able to better capture the nature of social networks~\cite{Wimmer2010}. Furthermore, we have used a connectivity kernel that: (a)~is symmetric and cannot identify whether there is a status order in society~\cite{Chan2004,Ball2013}, and (b)~only depends on differences between individuals. For example, young men tend to have more social contacts than young women, and older women have more social contacts than older men~\cite{Bhattacharya2016}---an observation that cannot be captured by a kernel of the form we have considered. The connectivity kernel could be refined by adding the demographic attributes of the seeds and nominees as features, capturing sociability and popularity, respectively. Furthermore, it should be determined whether the number of ``intervening opportunities''~\cite{Stouffer1940}, absolute distance in Blau space, or a hybrid thereof are most predictive of tie probability. Ultimately, learning a connectivity kernel without a pre-specified parametric form should be considered~\cite{Frolich2006} because they can better capture complex patterns, such as interactions between different demographic attributes. We note that, irrespective of the choice of connectivity kernel, the interpretable segregation statistics considered here remain valid and useful.

@@

We would like to thank Sahil Loomba and two anonymous referees for useful feedback on the manuscript and acknowledge support from the grant EP/N014529/1.

\printbibliography
